Cooperative Kernels: GPU Multitasking for Blocking Algorithms

### 2017-07-06 write-ups/FSE/paper.tex
Various.

\mytablong\ref{tab:applications} gives an overview of the 7 irregular
algorithms that we ported to cooperative kernels. Among them, 5 are
graph algorithms, based on the Pannotia~\cite{Pannotia} and
Lonestar~\cite{BNP12} GPU application suites, using global barriers.
We indicate how many of the original number of barriers are changed to
resizing barriers (all of them), and how many variables need to be
transmitted.  The remaining two algorithms are work stealing
applications: each required the addition of $\offerfork$ and
$\offerkill$ at the start of the main loop, and no variables needed to
be transmitted (similar to example shown in \mysec\ref{sec:programmingguidelines}). 
% For all these examples, the porting to cooperative
%kernels was similarly straightforward to the porting illustrated in
%\mysec\ref{sec:programmingguidelines}. 
Most graph applications come with 2 different data sets as input,
leading to 11 application/input pairs in total.

@@

\TSAdded{Our prototype implementation (\mysec\ref{sec:implementation}) requires
two optional features of OpenCL 2.0: SVM fine-grained buffers and SVM
atomics. Out of the GPUs available to us (ARM Mali, AMD, \nvidia, and
Intel), only Intel GPUs provided robust support of these features.}

@@

%Among the GPUs available to us, four met our requirements: Intel HD
%520, HD 5500 and Iris 6100, and AMD Radeon R7.  However, AMD's Linux
%drivers for OpenCL suffer from a known defect whereby long-running
%kernels lead to unkillable defunct
%processes~\cite{amd-defunct-process}.
%
%and simultaneously obtained alarming results on this platform when
%using SVM atomics: modifying a kernel to include a simple host/device
%handshake through SVM slowed down kernel execution by an order of
%magnitude in some cases.  This behaviour, combined with the defunct
%processes issue, lead us to suspect that the support we require for
%cooperative kernels is not yet mature enough within AMD's drivers,
%which we plan to discuss with engineers at AMD.
%
\TSAdded{We thus ran our experiments on the three Intel GPUs: HD 520, HD 5500
and Iris 6100. The results were similar across the GPUs, so for
conciseness, we report only on the Iris 6100 GPU (driver
20.19.15.4463) with a host CPU i3-5157U. The Iris has a reported 47
compute units. Results for the other Intel GPUs are available in
supplementary material, found on the author's website.}

@@

We devised the following to determine OpenCL workloads that simulate
the computational intensity of various graphics rendering workloads.
%
We designed a synthetic kernel that occupies all workgroups of a GPU
for a parameterised time period $t$, invoked in an infinite loop by a
host application.  We then searched for a maximum value for $t$ that
allowed the synthetic kernel to execute without having an observable
impact on graphics rendering.  Using the computed value, we ran the
host application for $X$ seconds, measuring the time $Y < X$ dedicated
to GPU execution during this period and the number of kernel launches
$n$ that were issued.  We used $X \geq 10$ in all experiments.  The
values $(X-Y)/n$ and $X/n$ estimate the average time spent using the
GPU to render the display between kernel calls (call this $E$) and the
period at which the OS requires the GPU for display rendering (call
this $P$), respectively.

@@

Our results are the following: $P=70ms$ and $E=3ms$ for light;
$P=40ms$, $E=3ms$ for medium; and $P=40ms$, $E=10ms$ for heavy. For
medium and heavy, the 40ms period coincides with the human persistence
of vision. The 3ms execution duration of both light and medium
configurations indicates that GPU computation is cheaper for basic
display rendering compared with more complex rendering.% that uses
%shaders.

@@

We show results for the three workloads listed in
\mysec~\ref{sec:sizingnoncoop}. The horizontal lines in the period
graph correspond to the goals of the workloads: the higher
(resp. lower) line corresponds to a period of 70ms (resp. 40ms) for
the light (resp. medium and heavy) workload.

Co-scheduling non-cooperative kernels that request a single workgroup
leads to almost no overhead, but the period is far too high to meet
the needs of any of our three workloads; e.g.\ a heavy workload
averages a period of 939 ms. As more workgroups are dedicated to
non-cooperative kernels, they execute quickly enough to be scheduled
at the expected period. For the light and medium workloads, a quarter
of the workgroups executing the non-cooperative kernel are able to
meet their goal period (70 and 40 ms resp.). However, this is not
sufficient to meet the goal for the heavy workload (giving a median
period of 104 ms). If half of the workgroups are allocated to the
non-cooperative kernel, the heavy workload achieves its goal period
(median of 40 ms).
%
Yet, as expected, allocating more non-cooperative workgroups increases
the overhead of the cooperative kernel.
%
Still, heavy workloads meet their period by allocating half
of the workgroups, incurring a slow down of less than
1.5$\times$ (median). Light and medium workloads meet their period
with only a small overhead; 1.04$\times$ and 1.08$\times$ median
slowdown respectively.

@@

Suppose a graphics workload is required to be scheduled with period
$P$ and duration $D$, and that a compute kernel requires time $C$ to
execute without interruption.  If we assume the cost of preemption is
negligible (e.g.\ Nvidia have reported preemption times of 0.1 ms for
Pascal~\cite{anandtech}), then the overhead associated with switching
between compute and graphics every $P$ time steps is $P/(P-D)$.

>>>>

\mytablong\ref{tab:applications} gives an overview of the 7 irregular
algorithms that we ported to cooperative kernels. Among them, 5 are
graph algorithms, based on the Pannotia~\cite{Pannotia} and
Lonestar~\cite{BNP12} GPU application suites, using global barriers.
We indicate how many of the original number of barriers are changed to
resizing barriers (all of them), and how many variables need to be
transmitted.  The remaining two algorithms are work stealing
applications: each required the addition of $\offerfork$ and
$\offerkill$ at the start of the main loop, and no variables needed to
be transmitted (similar to example discussed in \mysec\ref{sec:programmingguidelines}). 
% For all these examples, the porting to cooperative
%kernels was similarly straightforward to the porting illustrated in
%\mysec\ref{sec:programmingguidelines}. 
Most graph applications come with 2 different data sets as input,
leading to 11 application/input pairs in total.

@@

\TSAdded{Our prototype implementation (\mysec\ref{sec:implementation}) requires
two optional features of OpenCL 2.0: SVM fine-grained buffers and SVM
atomics. Out of the GPUs available to us, from ARM, AMD, \nvidia, and
Intel, only Intel GPUs provided robust support of these features.}

@@

%Among the GPUs available to us, four met our requirements: Intel HD
%520, HD 5500 and Iris 6100, and AMD Radeon R7.  However, AMD's Linux
%drivers for OpenCL suffer from a known defect whereby long-running
%kernels lead to unkillable defunct
%processes~\cite{amd-defunct-process}.
%
%and simultaneously obtained alarming results on this platform when
%using SVM atomics: modifying a kernel to include a simple host/device
%handshake through SVM slowed down kernel execution by an order of
%magnitude in some cases.  This behaviour, combined with the defunct
%processes issue, lead us to suspect that the support we require for
%cooperative kernels is not yet mature enough within AMD's drivers,
%which we plan to discuss with engineers at AMD.
%
\TSAdded{We thus ran our experiments on three Intel GPUs: HD 520, HD 5500
and Iris 6100. The results were similar across the GPUs, so for
conciseness, we report only on the Iris 6100 GPU (driver
20.19.15.4463) with a host CPU i3-5157U. The Iris has a reported 47
compute units. Results for the other Intel GPUs are available in
supplementary material, found on the author's website.}

@@

We devised the following method to determine OpenCL workloads that simulate
the computational intensity of various graphics rendering workloads.
%
We designed a synthetic kernel that occupies all workgroups of a GPU
for a parameterised time period $t$, invoked in an infinite loop by a
host application.  We then searched for a maximum value for $t$ that
allowed the synthetic kernel to execute without having an observable
impact on graphics rendering.  Using the computed value, we ran the
host application for $X$ seconds, measuring the time $Y < X$ dedicated
to GPU execution during this period and the number of kernel launches
$n$ that were issued.  We used $X \geq 10$ in all experiments.  The
values $(X-Y)/n$ and $X/n$ estimate the average time spent using the
GPU to render the display between kernel calls (call this $E$) and the
period at which the OS requires the GPU for display rendering (call
this $P$), respectively.

@@

Our results are the following: $P=70\mathit{ms}$ and $E=3\mathit{ms}$ for light;
$P=40\mathit{ms}$, $E=3\mathit{ms}$ for medium; and $P=40\mathit{ms}$, $E=10\mathit{ms}$ for heavy. For
medium and heavy, the $40\mathit{ms}$ period coincides with the human persistence
of vision. The $3\mathit{ms}$ execution duration of both light and medium
configurations indicates that GPU computation is cheaper for basic
display rendering compared with more complex rendering.% that uses
%shaders.

@@

We show results for the three workloads listed in
\mysec~\ref{sec:sizingnoncoop}. The horizontal lines in the period
graph correspond to the goals of the workloads: the higher
(resp. lower) line corresponds to a period of $70\mathit{ms}$ (resp.\ $40\mathit{ms}$) for
the light (resp. medium and heavy) workload.

Co-scheduling non-cooperative kernels that request a single workgroup
leads to almost no overhead, but the period is far too high to meet
the needs of any of our three workloads; e.g.\ a heavy workload
averages a period of 939 $\mathit{ms}$. As more workgroups are dedicated to
non-cooperative kernels, they execute quickly enough to be scheduled
at the expected period. For the light and medium workloads, a quarter
of the workgroups executing the non-cooperative kernel are able to
meet their goal period (70 and 40 $\mathit{ms}$ resp.). However, this is not
sufficient to meet the goal for the heavy workload (giving a median
period of 104 $\mathit{ms}$). If half of the workgroups are allocated to the
non-cooperative kernel, the heavy workload achieves its goal period
(median of 40 $\mathit{ms}$).
%
Yet, as expected, allocating more non-cooperative workgroups increases
the overhead of the cooperative kernel.
%
Still, heavy workloads meet their period by allocating half
of the workgroups, incurring a slow down of less than
1.5$\times$ (median). Light and medium workloads meet their period
with only a small overhead; 1.04$\times$ and 1.08$\times$ median
slowdown respectively.

@@

Suppose a graphics workload is required to be scheduled with period
$P$ and duration $D$, and that a compute kernel requires time $C$ to
execute without interruption.  If we assume the cost of preemption is
negligible (e.g.\ Nvidia have reported preemption times of 0.1 $\mathit{ms}$ for
Pascal~\cite{anandtech}), then the overhead associated with switching
between compute and graphics every $P$ time steps is $P/(P-D)$.
